---
title: "Getting started with biomimic"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with biomimic}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 96
)
```

## The problem

The standard way to analyse an antibodyome experiment is borrowed from
transcriptomics: test each protein separately for a difference between groups,
then correct for testing thousands of hypotheses. It works when a few proteins
have large individual effects. It struggles when the signal is spread across
many proteins that move together, because the multiple-testing burden grows
with every protein measured and marginal testing throws away the correlation
structure that carries the biology.

A MIMIC model asks a different question. Observed proteins are indicators of a
latent factor — an unobserved immunological state — and the clinical group acts
on that factor as a regressor. Instead of "which proteins differ?", the model
answers "does the latent state differ between groups?". One coefficient, one
test, no multiple-testing correction.

The catch is that structural equation modelling software wants you to name the
indicators. Choosing 15 out of 8,000 is not something it does, and specifying
and diagnosing the model assumes SEM training. `biomimic` handles both.

```{r setup}
library(biomimic)
```

## The example data

`neuro_antibodyome` is a subset of the ProtoArray data analysed in the paper,
for the Alzheimer's disease versus control comparison: 137 samples (46 cases,
91 controls) and 400 autoantibody features, centred and ready to use.

```{r data}
data(neuro_antibodyome)
str(neuro_antibodyome, max.level = 1)
```

The covariate matrix carries an intercept, the group indicator, and age and sex
as adjustment covariates.

```{r covariates}
head(neuro_antibodyome$X, 3)
```

## Running the pipeline

```{r fit}
fit <- biomimic(neuro_antibodyome$proteins,
                neuro_antibodyome$X,
                K = 1,
                k = 15)
fit
```

That single call did three things.

**Stage 1** fitted a variational Bayes factor model with ARD priors to all 400
features and ranked them by posterior signal-to-noise ratio. ARD shrinks
irrelevant loadings toward zero, so the ranking reflects contribution to the
shared latent factor rather than marginal effect size.

**Stage 2** kept the top 15.

**Stage 3** generated `lavaan` syntax for a MIMIC model on those 15 indicators
and fitted it. You never wrote a line of SEM syntax, but the model is a
standard one and the full `lavaan` object is available via `fit$lavaan_fit`.

## Reading the results

```{r summary}
summary(fit)
```

The line to read is the group effect on the latent pathway. It is negative and
strongly significant: the latent immunological state is lower in AD than in
controls, adjusted for age and sex. Age and sex have their own effects on the
same factor, which is exactly why they belong in the model.

Below that, the loadings show how each protein relates to the pathway. They are
uniformly high, which is what selection by signal-to-noise ratio is supposed to
produce — a coherent panel rather than a list of individually striking hits.

## A note on fit indices

The RMSEA above is flagged as poor. That deserves context rather than alarm.
The conventional cutoffs come from simulations with small, theory-driven models
and correctly specified indicators. They degrade as the number of indicators
grows, and high unique variances — expected with heterogeneous proteomic
indicators — inflate RMSEA and deflate CFI even for correctly specified models.
With a data-driven indicator set and an exploratory latent factor, treat
absolute cutoffs with caution and lean on SRMR and CFI alongside them.

```{r gof}
goodness_of_fit(fit)
```

## Screening on its own

You do not have to run the whole pipeline. Stage 1 is useful by itself as a
screening tool:

```{r screening}
vb <- vb_ard(neuro_antibodyome$proteins, neuro_antibodyome$X, K = 1)
head(rank_variables(vb), 10)
```

The `use_gamma` argument turns on the direct-effect extension, which places ARD
priors on covariate-to-indicator paths as well. It flags proteins whose
association with the covariate bypasses the latent factor — a step normally
done by hand from modification indices.

## Measurement invariance

If a latent difference between groups is to mean anything, the measurement
model has to behave the same way in both. `test_invariance()` fits the
configural, metric and scalar models and compares them.

```{r invariance, eval = FALSE}
test_invariance(fit$selection, group_var = "group")
```

Metric invariance means the loadings are equal across groups; scalar invariance
adds equal intercepts. Without them, a group difference on the factor could be
an artefact of the indicators behaving differently in each group.

## Plots

```{r plot-screening}
plot_screening(fit)
```

```{r plot-loadings}
plot_loadings(fit)
```

```{r plot-effects}
plot_structural_effects(fit)
```

## Where to go next

* `?vb_ard` — the screening model, its priors and the Gamma extension.
* `?build_lavaan_model` — the four model types and the syntax each produces.
* `?fit_sem` — estimator choice and Stage 3 options.
* `?network_analysis` — treating the selected panel as a co-expression module.
* `?simulate_mimic` — simulate data with known structure to evaluate the method.
